\documentclass[11pt,letterpaper]{article}
\setlength{\headheight}{14pt}

\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[margin=1in]{geometry}
\usepackage{wrapfig}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage[justification=centering]{caption}
\usepackage[subnum]{cases}
\setlength{\parindent}{0pt}
\setlength{\parskip}{1em}
\usepackage{empheq}
\usepackage{relsize}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{csquotes}
\usepackage{comment}
\graphicspath{ {./figures/} } % figures (images) go into here
\usepackage[
backend=biber,
style=ieee, % IEEE ref format
]{biblatex}
\addbibresource{bibliography.bib} %Imports bibliography file

\title{Cache Optimization: Are Our Algorithms Realistic?}
\date{\today}
\author{Logan Decock, Juan Rempel}

\begin{document}

\maketitle

\section{Abstract}
The speed of modern processors has caused memory latency to dominate processing time~\cite{demaine_lecture}. Usually, performance is measured through the lens of the word RAM model, ignoring the cost of fetching data across cache levels.
The cache-oblivious model is a cache efficient model of theoretical computation. Analysis under this model measures only the number of transfers between two adjacent cache levels without knowing block or total cache sizes. As a result, algorithms optimized for the cache-oblivious model are efficient between any two given levels of the memory hierarchy. In our paper we will explore the conditions in which the real world performance of select algorithms benefit from optimization within the cache-oblivious model over the word RAM model. We will discover and analyse data size and bus latency thresholds for which cache-oblivious optimized versions of said algorithms outperform their word RAM optimized counterparts.
% We will find thresholds where select algorithms outperform their counterparts in the other model, finding the conditions under which the cache-oblivious optimized algorithm should be chosen for optimal real-world performance.

\section{Motivation}
% Our paper will evaluate the accuracy of the cache-oblivious model opposed to the word RAM model in predicting real world algorithm performance. We will consider analogous algorithms and data structures optimized for the cache-oblivious model and the word RAM model. We will analyse these algorithm's asymptotic cost in both models and then apply these algorithms to synthetic datasets to measure their real world performance. 

With data fetching being a real cost, orders of magnitude greater than that of a single instruction, we think performance gains can be achieved through algorithm design that measure factors beyond instruction count, namely cache misses. The cache-oblivious model was created as solution to this problem. Unlike the word RAM model, the cache-oblivious model counts only the number of memory transfers between any two levels of the memory hierarchy, completely ignoring instruction count.

Anecdotally, we have found algorithms optimized for the cache-oblivious model tend to have larger constants hidden in their asymptotic cost compared to those optimized for the word RAM model. These hidden constants result in worse runtime performance on small datasets. We will explore how cache-oblivious algorithms scale as our dataset grows. When data gets large enough, a miss at one level of the memory hierarchy can require a request over the network, incurring costs on the scale of milliseconds.

With these massive costs in mind, we believe the cache-oblivious model has the opportunity to supersede the word RAM model for designing algorithms that process massive datasets. Our paper intends to give developers tools to analytically choose algorithms from the optimal model for their dataset. Such a tool is especially relevant today with the prevalence of big data with multi petabyte datasets, much too large to ignore in real-world performance analysis in a strictly word RAM model. 

% We hope that in our paper to prove through experiment that the cache-oblivious model performs better than the word RAM model on large datasets as well as find the dataset size tipping point at which the cache-oblivious becomes the superior model in which to optimize the algorithm.

\section{Format}
Our project will mostly consist of a performance comparison paper. We will find a few algorithms that we wish to examine, having already found the cache-oblivious counterpart to merge sort being funnelsort~\cite{demaine}. To test and analyze the algorithms, we will run them through simulations where we simulate the cache misses. 

Within our research, we will be relying heavily on Demaine's paper as the foundation for our understanding of cache-oblivious algorithms~\cite{demaine}. Since our research is not meant to know specific hardware specifications, we will ignore other similar models such as the cache-aware or external memory models. We will start with our algorithm selection by looking at two papers by Brodal et al., in which are described a \emph{lazy funnelsort} and \emph{binary tree}~\cite{brodal_bst}~\cite{brodal_lazyfunnel}. Since we'll be implementing the code for the funnelsort, we may also need to reference the original \emph{funnelsort} as introduced by Frigo et al.~\cite{frigo}.

\section{Timeline}
\begin{itemize}
    \item \textbf{March 10:} select 1--2 other algorithms to analyze
    \item \textbf{March 10:} understand the selected agorithms in their respective theoretical models
    \item \textbf{March 17:} program completed for our experimentation, proceed to experimentation
    \item \textbf{March 22:} experimentation and analysis completed; start presentation slides
    \item \textbf{March 24:} presentation slides completed, presentation rehearsed; start paper
    \item \textbf{April 7:} paper completed 
\end{itemize}

% \clearpage
% We just add things to the bibliography.bib file, and if we cite it, then it will get printed. Uncited items will not be printed.
\printbibliography%

\end{document}
